Legion: Expressing Locality and Independence with Logical Regions

### 2012-05-05 doc/sc2012/experiments.tex
updates

Our second experiment is a port of the \emph{fluidanimate} benchmark from the PARSEC suite\cite{bienia11benchmarking},
which does a particle-based simulation of an incompressible fluid.  Each particle interacts only with other nearby
particles. The benchmark divides the space in which the fluid can move into a three-dimensional array of cells 
such that the range of interaction is limited to just the
cells adjacent (including diagonals) to the one a particle resides in.  The application divides the array 
into {\em grids} and assigns each one to 
a thread.  Per cell locking is used to safely access particles in cells that lie
on the edge of a grid.  This fine-grained locking scheme along with the assumption of a shared address space leads to
good scaling in a multi-core processor, but rules out any attempt run beyond a single node.

To extend the scaling further, our port uses region partitioning to divide the array into grids in a similar
way, but handles the interaction between grids in a way that doesn't rely on access to shared memory.
Instead, the Legion version of the benchmark creates explicit ghost copies of cells on the boundaries of grids
and uses those ghost cells to exchange information between the grids.

The mapper for the particle simulation is very simple - it maps one grid's tasks onto each processor and
maps all regions for that grid (both the internal and the ghost regions) into the system memory used by that processor.
The exchange of the ghost cell data between processors is handled by the Legion runtime as a ghost cell region
gets alternately mapped in two different memories.

Figure~\ref{fig:fluid_single} compares the performance of the Legion implementation against the PARSEC version,
using a relatively small problem (300K particles on a 15x21x15 array of cells).  Between 1 and 4 threads, the
results for Sapling are nearly indistinguishable, indicating 
neither the Legion runtime nor the restructuring of the implementation to allow multi-node scaling impose
any significant overhead.  
%It's possible that the use of explicit ghost cells rather than fine-grained sharing of cache lines might be a net win.  
At 8 threads and above, the performance begins to vary.
Both the Legion and PARSEC version on Viz flatten out as they over-subscribe the 12 physical cores.  On Sapling,
which has HyperThreading enabled, deviations from linear begin sooner as the operating system's scheduler's
thread placement choices begin to matter.

To measure scaling beyond a single node, three different problem sizes were run for each of the three systems,
 with the results plotted in Figure~\ref{fig:fluid_multi}.  
 For the smallest problem (300K particles), we observe a 20\% speedup when going to 1
node to 2 (a total of 16 threads), but begin slowing down beyond that due to communication overhead - at 4 nodes
there are twice as many ghost cells as interior grid cells.  The larger problem sizes (2.4M and 19M
particles) do much better, with scaling of up to 5.4x when going from 1 node up to 16 because of a lower 
communication-to-computation ratio.

@@

\subsection{Adaptive Mesh Refinement}
\label{subsec:exp_amr}
Our final application is based on the third heat equation example from the Berkeley Labs 
BoxLib project \cite{BoxLib}.  This application is a three-level adaptive-mesh-refinement (AMR)
code that computes a first order stencil on a 2D mesh of cells.  The simulation iterates for many time
steps.  Updating the simulation for one time step consists of three phases.  In the first phase,
the boundary cells around a box at a refined level linearly interpolate their values from the nearby cells at the
next coarsest level.  The second phase performs the stencil computation on each cell in every level.
In the third phase, cells at a coarser level that have been refined are restricted to the average
of the cells that they physically contain at the next finest level of refinement.

Achieving high-performance on this application is particularly challenging for several reasons.  
First, the application has a very high communication-to-computation ratio which, for a fixed problem
size begins as being memory bound and with increasing node count becomes network bound as the
perimeter-to-area ratio of cell grids increases.  Second, when choosing how to partition cells into grids, the
programmer must consider the locality both between cells in the same level, as well as for cells
across levels.  In the case of cross-level cell dependences, mapping decisions must be made at runtime as the
location of refinements are determined based on application input.  Finally, this application has
parallelism both between tasks running at the same level and tasks running across levels, leading
to complicated input-dependent data dependences.

BoxLib's implementation of this benchmark partitions cells within the same level into a number of grids
based on the number of nodes in the machine and distributes one grid from each level to each node.  This
optimizes for memory bandwidth and load balance, but does not take into account cross-level locality 
that exists between grids from different levels of refinement.  Furthermore, BoxLib does not block grids
into sub-grids to take advantage of intra-grid locality.

Our Legion implementation performs two optimizations that allow us to achieve better performance than BoxLib.
First, for each level of refinement we recursively partition the logical region of cells based on 
the number of nodes in the machine and the size of the L3 cache and L2 caches.  Legion allows us to describe 
locality for many levels of the memory hierarchy instead of just at the node-level.

Our second optimization takes advantage of the cross-level locality present in the application.  We implemented
an application specific mapper that dynamically discovers the relationships between
grids at different levels of refinement.  The mapper dynamically performs intersection tests between logical regions
containing grids of different refinement levels to determine whether they overlap.  
If grids from different levels overlap, the mapper
ensures that they are always placed on the same node in the machine to leverage cross-level locality.  The mapper
memoizes the results of these intersection tests to amortize their cost.  The mapper also dynamically
load balances by distributing unconstrained grids from the coarsest level onto under-loaded nodes.

We compared our Legion implementation against BoxLib on three different problem sizes with a fixed number
of cells per level of refinement, but with randomly chosen refinements locations.  BoxLib also supports OpenMP
and we took their best performance from using 1, 2, 4, or 8 threads per node.  Our Legion implementation
always uses one thread per node to illustrate that in this application locality is significantly more important 
than parallelism.  The results from our experiments on all three clusters can be seen in Figure~\ref{fig:amr_total}.

On just one node, blocking for caches using Legion achieves up to 2.6X speedup over BoxLib.  As the node count 
increases, the ability of our mapper to leverage cross-level locality allows us to further increase
our performance advantage to 5.4X over BoxLib by reducing the total communication costs.

As the node count increases the AMR code becomes highly dependent on interconnect performance.  BoxLib performs much better
on Keeneland than it does on our viz cluster due to the better interconnect.  This can also be seen as we scale
to a larger number of nodes and BoxLib begins to catch up as can be seen in Figure~\ref{fig:amr_keeneland}.  
This occurs because our intra-level ghost-cell exchange algorithm relies on GASNet memory to exchange ghost cells which 
involves a linear increase in network traffic with the number of nodes.  BoxLib instead uses direct
node-to-node exchanges of ghost cells similar to our fluid application.  A future implementation of our AMR code 
will employ a similar ghost cell exchange algorithm to improve scalability.

>>>>

Our second experiment is a port of the \emph{fluidanimate} benchmark
from the PARSEC suite\cite{bienia11benchmarking}, which does a
particle-based simulation of an incompressible fluid.  Each particle
interacts only with other nearby particles. The benchmark divides the
space in which the fluid can move into a three-dimensional array of
cells such that the range of interaction is limited to just the cells
adjacent (including diagonals) to the one a particle resides in.  The
application divides the array into {\em grids} and assigns each grid to
a thread.  Per-cell locking safely accesses particles in
cells that lie on the edge of a grid.  This fine-grained locking
scheme along with the assumption of a shared address space gives
good scaling in a multi-core processor, but rules out any attempt to run
beyond a single node.

To extend the scaling further, our port uses region partitioning to
divide the array into grids in a similar way, but avoids relying
on shared memory to handle interactions between grids.
Instead, the Legion version creates
explicit ghost copies of cells on grid boundaries and uses
those ghost cells to exchange information between the grids.

The particle simulation's mapper is very simple: it maps one grid's
tasks onto each processor and maps all that grid's regions (both
internal and ghost regions) into that processor's system memory.  The
exchange of ghost cell data between processors is handled by the
Legion runtime as a ghost cell region is alternately mapped to two
different memories.

Figure~\ref{fig:fluid_single} compares the performance of the Legion
implementation against the PARSEC version, using a relatively small
problem (300K particles on a 15x21x15 array of cells).  Between 1 and
4 threads, the results for Sapling are nearly indistinguishable,
indicating neither the Legion runtime nor the restructuring of the
implementation to allow multi-node scaling impose any significant
overhead.
%It's possible that the use of explicit ghost cells rather than fine-grained sharing of cache lines might be a net win.  
At 8 threads and above, performance begins to vary.  Both the Legion
and PARSEC versions on Viz flatten out as they over-subscribe the 12
physical cores.  On Sapling, which has HyperThreading enabled,
deviations from linear begin sooner as the operating system's
thread placement choices begin to matter.

To measure scaling beyond a single node, three different problem sizes
were run for each of the three systems (Figure~\ref{fig:fluid_multi}).
For the smallest problem (300K particles), we observe a 20\% speedup
from 1 to 2 nodes (16 threads total), but slow
down beyond that due to communication overhead---at 4 nodes there are
twice as many ghost cells as interior grid cells.  The larger problem
sizes (2.4M and 19M particles) do much better, with scaling of up to
5.4x when going from 1 node up to 16 because of a lower
communication-to-computation ratio.

@@

\subsection{Adaptive Mesh Refinement}
\label{subsec:exp_amr}
Our final application is based on the third heat equation example from
the Berkeley Labs BoxLib project \cite{BoxLib}.  This application is a
three-level adaptive-mesh-refinement (AMR) code that computes a first
order stencil on a 2D mesh of cells.  
%The simulation iterates for many time steps.  
Updating the simulation for one time step consists of
three phases.  In the first phase, the boundary cells around a box at
a refined level linearly interpolate their values from the nearby
cells at the next coarsest level.  The second phase performs the
stencil computation on each cell in every level.  In the third phase,
cells at a coarser level that have been refined are restricted to the
average of the cells that they physically contain at the next finest
level of refinement.

Achieving high-performance on this application is particularly
challenging for several reasons.  First, the application has a very
high communication-to-computation ratio which, for a fixed problem
size, begins as being memory bound and with increasing node count
becomes network bound as the perimeter-to-area ratio of cell grids
increases.  Second, when choosing how to partition cells into grids,
the programmer must consider the locality between cells within a
level as well as across levels.  For
cross-level cell dependences, mapping decisions must be made at
runtime as the location of refinements are dynamically determined.
Finally, this application has parallelism both
between tasks running at the same level and tasks running across
levels, leading to complicated input-dependent data dependences.

BoxLib's implementation partitions cells within a level 
into a number of grids based on the number of nodes in the
machine and distributes one grid from each level to each node.  This
optimizes for memory bandwidth and load balance, but does not 
exploit cross-level locality between grids from
different levels of refinement.  Furthermore, BoxLib does not block
grids into sub-grids to take advantage of intra-grid locality.

Our Legion implementation performs two optimizations that allow us to
outperform BoxLib.  First, for each level of
refinement we recursively partition the logical region of cells based
on the number of nodes in the machine and the sizes of the L2 and L3 caches.
%Legion allows us to describe locality for many levels of
%the memory hierarchy instead of just at the node-level.
Our second optimization takes advantage of the cross-level locality.
We wrote an application-specific mapper that dynamically discovers
relationships between grids at different levels of refinement.  The
mapper dynamically performs intersection tests between logical regions
containing grids of different refinement levels.  If the mapper
discovers overlaps between grids from different levels, the mapper
places them on the same node in the machine.  The
mapper memoizes the intersection tests to amortize their cost.  The
mapper also dynamically load balances by distributing unconstrained
grids from the coarsest level onto under-loaded nodes.

We compared our Legion implementation against BoxLib on three
different problem sizes with a fixed number of cells per level of
refinement, but with randomly chosen refinement locations.  BoxLib
also supports OpenMP and we took their best performance from using 1,
2, 4, or 8 threads per node.  Our Legion implementation always uses
one thread per node to illustrate that in this application locality is
significantly more important than parallelism.

Figure~\ref{fig:amr_total} gives the results.
On just one node, blocking for caches using Legion achieves up to 2.6X
speedup over BoxLib.  As node count increases, the mapper's
ability to exploit cross-level locality further increases
the performance advantage to 5.4X by reducing the total
communication costs.

As the node count increases the AMR code becomes highly dependent on
interconnect performance.  BoxLib performs much better on Keeneland
than on Viz due to the better interconnect.  
At higher node counts BoxLib
begins to catch up (see Figure~\ref{fig:amr_keeneland})
because our intra-level ghost-cell exchange 
uses GASNet memory to exchange ghost cells, requiring a
linear increase in network traffic with the number of nodes.  BoxLib
uses direct node-to-node exchanges of ghost cells, similar to
our fluid application.  A future implementation of our AMR code will
employ a similar ghost cell exchange algorithm to improve scalability.
